Measurement and Probability in Fuzzy Quantum Theories

### 2016-11-27 proposal/prop.tex
p

\item Deutsch-Jozsa: In a follow-up paper~\cite{DQT2014}, we proposed
  a variant of quantum mechanics in which the underlying field of
  complex numbers is replaced by special finite fields with enough
  structure to model Hermitian products that \emph{locally} behave
  like inner products. The trick is to recognize that, as the
  characteristic of the finite field increases, larger and larger
  regions of the finite field become totally ordered, thereby
  restoring, with some restrictions, all the usual properties of the
  inner product needed for conventional quantum mechanics. Using this
  theory to analyze the Deutsch -Jozsa algorithm reveals that the
  characteristic of the finite field plays an essential role: as the
  size of the finite field grows, so does the numerical range of the
  intermediate and final results of the algorithms being implemented.
  The theory naturally recovers a deterministic measure of the
  intrinsic resources required for a given level of complexity; this
  measure is normally completely hidden in computations with real
  numbers, and explicitly exposing it is one of the significant
  achievements of our discrete field analysis of quantum computation.
  This solves the conundrum that the conventional Deutsch-Jozsa
  algorithm mysteriously continues to work for larger and larger input
  functions without any apparent increase in resources.  Our analysis
  of this problem reveals that as the size of the input increases, it
  is necessary to increase the size of the finite field and hence the
  size of the underlying available numeric coefficients.  This
  observation does not fully explain the power of quantum computing
  over classical computing, but at least it explains that some of the
  power of quantum computing depends on increasingly larger precision
  in the underlying field of numbers.

@@

\noindent (A) {\em Integration of research and education:} Courses on
advanced computational models and quantum information will be further
developed at Indiana University to communicate the basic concepts and
tools discussed in this proposal. Dr. Ortiz taught courses on
``Quantum information and Computation'' and Dr. Sabry has taught
courses on ``Quantum Computing'' for graduate students.  These courses
were highly appreciated by the students as noted by their evaluations
in making the material extremely accessible and interesting.  The PIs
plan to engage high school students or undergraduates in the
implementation of some of the ideas and the exploration of finite
geometry.  The PIs have a long standing commitment to K-12 science and
math education (see also (C) below). The collaborative group efforts
will provide students, as well as the PIs with a unique learning
experience. The broadened perspectives gained through those efforts
will contribute to the PIs effectiveness as teachers and mentors. They
also will enable to better prepare students and postdocs for
productive careers in an increasingly competitive global
economy. Students and postdocs are encouraged to pursue collaborative
research leading to a broader scientific program. This is highlighted
by the PIs students' close contact with LANL, the Perimeter Institute,
and the Institute for Quantum Computing at Waterloo.

@@

Over the past several years, the PIs, together with other IU faculty
members and students, has formed a group whose main interest is the
study of paradigms of computing that either arise in nature or are
inspired by natural phenomena. The PI is especially interested in how
the laws of physics can inspire and guide our abstract computational
models, as well as exposing new insights into the computational
framework behind the natural world and the computability of physical
theory. Our group has been active in several areas related to the
theory behind quantum computation. Particular areas include:
mathematical foundations of quantum computation and
complexity~\cite{AGVS2007,VAS2006}, adiabatic quantum computing,
quantum logic, classical and quantum reversible programming
languages~\cite{Sabry:2003:MQC:871895.871900}, analog computation and
its limits, and optical computing. During the last year, our
discussions have revolved around formulating variants of quantum
mechanics that are computable and that are as close to possible to
classical computing. Several preliminary results have emerged from our
collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of programming
language research.

Applying within the university.. NSF funding would help us to get
critical mass; new engineering program; credibility importance
university will fund; new lab entanglement

There was no recent NSF support for Dr. Ortiz. In the past five years,
Prof. Sabry received one NSF grant (award number 1217454) titled
\emph{Information Effects} for the amount of \$275,319.00 with start
and end dates of 1 September 2012 -- 31 August 2016.

% (c) a summary of the results of the completed work, including
% accomplishments, supported by the award. The results must be
% separately described under two distinct headings, Intellectual Merit
% and Broader Impacts;

@@

\paragraph*{Intellectual Merit.}

\paragraph*{Broader Impacts.}

>>>>

\item Deutsch-Jozsa: In a follow-up paper~\cite{DQT2014}, we proposed
  a variant of quantum mechanics in which the underlying field of
  complex numbers is replaced by special finite fields with enough
  structure to model Hermitian products that \emph{locally} behave
  like inner products. The trick is to recognize that, as the
  characteristic of the finite field increases, larger and larger
  regions of the finite field become totally ordered, thereby
  restoring, with some restrictions, all the usual properties of the
  inner product needed for conventional quantum mechanics. Using this
  theory to analyze the Deutsch-Jozsa algorithm reveals that the
  characteristic of the finite field plays an essential role: as the
  size of the finite field grows, so does the numerical range of the
  intermediate and final results of the algorithms being implemented.
  The theory naturally recovers a deterministic measure of the
  intrinsic resources required for a given level of complexity; this
  measure is normally completely hidden in computations with real
  numbers, and explicitly exposing it is one of the significant
  achievements of our discrete field analysis of quantum computation.
  This solves the conundrum that the conventional Deutsch-Jozsa
  algorithm mysteriously continues to work for larger and larger input
  functions without any apparent increase in resources.  Our analysis
  of this problem reveals that as the size of the input increases, it
  is necessary to increase the size of the finite field and hence the
  size of the underlying available numeric coefficients.  This
  observation does not fully explain the power of quantum computing
  over classical computing, but at least it explains that some of the
  power of quantum computing depends on increasingly larger precision
  in the underlying field of numbers.

@@

\noindent (A) {\em Integration of research and education:} Courses on
advanced computational models and quantum information will be further
developed at Indiana University to communicate the basic concepts and
tools discussed in this proposal. Dr. Ortiz taught courses on
``Quantum information and Computation'' and Dr. Sabry has taught
courses on ``Quantum Computing'' for graduate students.  These courses
were highly appreciated by the students as noted by their evaluations
in making the material extremely accessible and interesting.  The PIs
plan to engage high school students or undergraduates in the
implementation of some of the ideas.  The PIs have a long standing
commitment to K-12 science and math education (see also (C)
below). The collaborative group efforts will provide students, as well
as the PIs with a unique learning experience. The broadened
perspectives gained through those efforts will contribute to the PIs
effectiveness as teachers and mentors. They also will enable to better
prepare students and postdocs for productive careers in an
increasingly competitive global economy. Students and postdocs are
encouraged to pursue collaborative research leading to a broader
scientific program. This is highlighted by the PIs students' close
contact with LANL, the Perimeter Institute, and the Institute for
Quantum Computing at Waterloo.

@@

\paragraph*{PI Team.} Over the past several years, the PIs, together
with other IU faculty members and students, have formed a group whose
main interest is the study of paradigms of computing that either arise
in nature or are inspired by natural phenomena. The PIs are especially
interested in how the laws of physics can inspire and guide our
abstract computational models, as well as exposing new insights into
the computational framework behind the natural world and the
computability of physical theory. Our group has been active in several
areas related to the theory behind quantum computation. Particular
areas include: mathematical foundations of quantum computation and
complexity~\cite{AGVS2007,VAS2006}, adiabatic quantum computing,
quantum logic, classical and quantum reversible programming
languages~\cite{Sabry:2003:MQC:871895.871900}, analog computation and
its limits, and optical computing. During the past several years, our
discussions have revolved around formulating variants of quantum
mechanics that are computable and that are as close to possible to
classical computing. Several preliminary results have emerged from our
collaboration~\cite{usat,finiteQC,geometry2013,DQT2014}, and the group
is very excited about the potential implications of those
results. This proposal brings forth a particular aspect of this
programme related to its applications to problem areas that have
traditionally been outside the conceptual framework of quantum
mechanics and programming language research. This proposal, if funded,
will give our group more recognition within the university which, at
this point, is committed to generously funding several ``emerging
areas of research''. With the establishment of a new Intelligent
Systems Engineering at Indiana University with a focus nanoscale
sensors and quantum engineering and the recent hires in Physics in
experimental atomic physics, ion-trap quantum simulation, computation,
and measurement, the PI group is in an excellent position to receive
additional funding from the University to reach a critical mass of
quantum information researchers.

\paragraph*{Prior Support.} There was no recent NSF support for
Dr. Ortiz. In the past five years, Prof. Sabry received one NSF grant
(award number 1217454) titled \emph{Information Effects} for the
amount of \$275,319.00 with start and end dates of 1 September 2012 --
31 August 2016.

\paragraph*{Intellectual Merit.} Computing is about processing
information. Physically speaking, information, like energy, is a
resource that obeys a conservation principle: information can never be
created nor destroyed; it is just transformed. Traditional models of
computation, however, have not embraced this conservation of
information as an explicit principle. Instead this principle appears,
after the fact, in applications such as information-flow security
which treats the amount of information released to the environment as
a possible security leak and such as differential privacy which
intentionally releases some random information to the user to confuse
potential attackers. This project has instead embraced the principle
of conservation of information as a key foundation for the design of
computational models. In one perspective, computation in these models
is constructed from reversible transformations that act on topological
spaces, deforming them, stretching them, and shrinking them but never
cutting nor glueing. The first main result of this project was the
design and implementation of a programming language which captures
these ideas and realizes them in the context of the new foundational
framework of Homotopy Type Theory. This framework combines ideas from
mathematics, logic, physics, and computation in a way that accounts
for a newly-recognized higher-dimensional structure of types. This
idea enabled us to not only represent programs as reversible
deformations but also to represent the reversible deformations
themselves as higher-level programs implementing reversible
optimizations of the underlying lower-level programs. In a second
breakthrough, viewing the higher-level programs as distinct proofs of
reversibility enabled the definition of types whose natural
cardinality is fractional. Having computations that manipulate
fractional types opens the door for a rich and novel class of
applications in which resource creation and resource use are
separated. This idea brings speedups and convenience in a way that is
similar to the speedup and convenience brought by the use of credit
cards to complete transactions without necessarily having the money at
hand: the monetary resource can be consumed by the merchant and only
later be reconciled by the buyer.

\paragraph*{Broader Impacts.} The idea of reversible computation and
its connection to quantum mechanics is at the heart of one of the
deepest mysteries in the universe. Any progress or theory that
explains physical principles computationally is likely to change our
worldview impacting everything from philosophy and religion to all
conventional sciences. On the development of researchers, the project
resulted in the graduation of one Ph.D. student with another one
expected to graduate in Spring 2017, the current supervision of two
additional Ph.D. students, the graduation of an MS student, and the
hosting of a visiting scholar.

\paragraph*{Publications.} Several papers were published to report on
the results. On the foundational side, we have the following papers:
\begin{itemize}
\item Computing with Semirings and Weak Rig Groupoids, European
  Symposium on Programming, 2016.
\item Superstructural Reversible Logic, International Workshop on
  Linearity, 2014.
\item Theseus: A High Level Language for Reversible
  Computing. Work-in-Progress report in the Conference on Reversible
  Computation, 2014.
\item Discrete Quantum Theories, J. Phys. A: Math. Theor. 2014.
\item Geometry of Discrete Quantum Computing, J. Phys. A:
  Math. Theor. 2013.
\item Isomorphic Interpreters from Logically Reversible Abstract
  Machines. Workshop on Reversible Computation 2012.
\end{itemize}
On the applications side, we have the following papers: 
\begin{itemize}
\item Copredication in Homotopy Type Theory: A homotopical approach
  for formal semantics of natural languages. Workshop on Natural
  Language and Computer Science, 2016.
\item Contract Monitoring Semantics as Patterns of
  Communication. International Conference on Functional Programming,
  2015.
\item Extensible Effects: An Alternative to Monad
  Transformers. Haskell Symposium, 2013.
\item Encoding Secure Information Flow with Restricted Delegation and
  Revocation in Haskell. Workshop on Functional Programming Concepts
  in Domain-Specific Languages, 2013.
\end{itemize}                                                                          
In addition to the publications, Dr. Sabry and collaborators maintain
a public repository with software related to the project. Dr. Sabry
was also invited to give talks at several venues, most notably:
\begin{itemize}
\item Colloquiumfest, University of Saskatchewan, Canada, 2014 
\item Workshop on Quantum Information and Foundations of Quantum
  Mechanics, University of British Columbia, Vancouver, Canada, 2013
\end{itemize}

@@
